---
title: "Mouse heart RNA-seq - ROO and ROO+omega3 diet analysis"
author: "del-Rio-Vazquez, Jose L."
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 4
    number_sections: true
    theme: flatly
    highlight: tango
    code_folding: show
    df_print: paged
---

```{r setup, include=FALSE}
# Global chunk options: suppress messages and warnings for a cleaner HTML report.
# All figures are centred at 300 dpi. A fixed random seed guarantees that
# stochastic steps (ggrepel label placement, fgsea permutations) are
# reproducible across runs.
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 8,
  fig.height = 5.5,
  dpi = 300
)
set.seed(42)
```

# Scope

This R Markdown analysis is adapted to **mouse heart RNA-seq** data from dietary intervention groups.

Included samples:

- `Control`: samples 9, 11 and 12.
- `ROO`: samples 5, 6, 7 and 8.
- `ROO_omega3`: samples 17, 18, 19 and 20.

The analysis intentionally excludes all other samples in the original count matrix. The expected input is a raw gene count matrix with a first column named `gene_id` or `Geneid`, and sample columns named as in the uploaded matrix, e.g. `sample_05`, `sample_06`, etc.

The workflow generates QC plots, PCA, DESeq2 differential expression tables, volcano plots, MA plots, contrast-level DEG heatmaps, contrast-level targeted cardiac-process heatmaps, gene-level boxplots, GSEA/FGSEA enrichments, contrast-level NES heatmaps and chord plots linking enriched pathways to leading-edge genes. In every heatmap, columns represent contrasts rather than individual samples. In chord plots, red indicates upregulation in the numerator of the contrast and blue indicates downregulation.

# Parameters

```{r params}
# --------------------------------------------------------------------------
# INPUT / OUTPUT PATHS
# The script looks for the count matrix in three locations in order:
#   1. Working directory
#   2. input_data/ subfolder
#   3. ../input_data/ (one level up)
# All outputs are written under OUT_DIR, organised into six subdirectories.
# --------------------------------------------------------------------------
COUNT_FILE <- "gene_count_matrix.csv"
if (!file.exists(COUNT_FILE)) COUNT_FILE <- file.path("input_data", "gene_count_matrix.csv")
if (!file.exists(COUNT_FILE)) COUNT_FILE <- file.path("..", "input_data", "gene_count_matrix.csv")

OUT_DIR <- "results_mouse_heart_diet"
DIRS <- list(
  qc = file.path(OUT_DIR, "01_qc"),
  de = file.path(OUT_DIR, "02_differential_expression"),
  enrichment = file.path(OUT_DIR, "03_enrichment"),
  targeted = file.path(OUT_DIR, "04_targeted_gene_sets"),
  boxplots = file.path(OUT_DIR, "05_gene_boxplots"),
  tables = file.path(OUT_DIR, "06_tables")
)
invisible(lapply(c(OUT_DIR, unlist(DIRS)), dir.create, recursive = TRUE, showWarnings = FALSE))

# --------------------------------------------------------------------------
# ANALYSIS THRESHOLDS
# MIN_COUNT / MIN_SAMPLES: a gene is retained if it has >= MIN_COUNT raw
#   reads in at least MIN_SAMPLES samples (low-expression filter).
# PADJ_CUTOFF / LFC_CUTOFF: Benjamini–Hochberg FDR and |log2FC| thresholds
#   used to call differentially expressed genes (DEGs).
# TOP_* constants control how many genes / terms are shown in each plot.
# --------------------------------------------------------------------------
MIN_COUNT <- 10
MIN_SAMPLES <- 3
PADJ_CUTOFF <- 0.05
LFC_CUTOFF <- 1
TOP_LABEL_GENES <- 12
TOP_HEATMAP_GENES <- 50
TOP_ENRICHMENT_TERMS <- 20
TOP_ENRICHMENT_HEATMAP_TERMS <- 40
```

# Packages and helper functions

```{r packages}
# --------------------------------------------------------------------------
# REQUIRED PACKAGES
# Bioconductor: DESeq2, SummarizedExperiment, AnnotationDbi, org.Mm.eg.db,
#               msigdbr, fgsea, ComplexHeatmap
# CRAN: tidyverse stack, ggplot2/ggrepel/patchwork, pheatmap, circlize,
#       RColorBrewer, scales, kableExtra
# The script stops with a clear error message if any package is missing.
# --------------------------------------------------------------------------
required_packages <- c(
  "DESeq2", "SummarizedExperiment", "AnnotationDbi", "org.Mm.eg.db",
  "dplyr", "tibble", "readr", "stringr", "tidyr", "purrr", "forcats",
  "ggplot2", "ggrepel", "scales", "RColorBrewer", "patchwork",
  "pheatmap", "ComplexHeatmap", "circlize", "grid", "kableExtra",
  "msigdbr", "fgsea"
)

missing_packages <- required_packages[!vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing_packages) > 0) {
  stop(
    "Missing required packages: ", paste(missing_packages, collapse = ", "),
    "\nInstall CRAN packages with install.packages() and Bioconductor packages with BiocManager::install()."
  )
}
suppressPackageStartupMessages(lapply(required_packages, library, character.only = TRUE))

# Explicit namespace binding to avoid conflicts between dplyr and other packages
# (e.g. MASS::select, stats::filter) that are silently loaded as dependencies.
select <- dplyr::select
filter <- dplyr::filter
mutate <- dplyr::mutate
arrange <- dplyr::arrange

# --------------------------------------------------------------------------
# HELPER FUNCTIONS
# heart_theme(): custom ggplot2 theme applied to all figures in this report.
# safe_name():  converts a contrast label into a filesystem-safe string
#               (replaces +, ω and non-alphanumeric characters with "_").
# save_plot():  saves a ggplot object as both SVG (vector) and PNG (raster).
# make_kable(): renders a data frame as a styled HTML table in the report.
# clean_pathway_name(): strips MSigDB prefixes and formats pathway names
#               for display in plots and heatmaps.
# --------------------------------------------------------------------------
heart_theme <- function(base_size = 12) {
  theme_classic(base_size = base_size) +
    theme(
      axis.text = element_text(color = "black"),
      plot.title = element_text(face = "bold", hjust = 0),
      plot.subtitle = element_text(color = "grey35", size = rel(0.9)),
      legend.title = element_text(face = "bold"),
      panel.grid.major = element_line(linewidth = 0.2, color = "grey92")
    )
}

safe_name <- function(x) {
  x |>
    stringr::str_replace_all("\\+", "plus") |>
    stringr::str_replace_all("ω", "omega") |>
    stringr::str_replace_all("[^A-Za-z0-9_]+", "_") |>
    stringr::str_replace_all("_+", "_") |>
    stringr::str_replace_all("_$", "")
}

save_plot <- function(plot, filename, width = 8, height = 5.5, dir = DIRS$de) {
  fn_svg <- file.path(dir, paste0(filename, ".svg"))
  fn_png <- file.path(dir, paste0(filename, ".png"))
  ggplot2::ggsave(fn_svg, plot = plot, width = width, height = height, dpi = 300)
  ggplot2::ggsave(fn_png, plot = plot, width = width, height = height, dpi = 300)
  invisible(c(fn_svg, fn_png))
}

make_kable <- function(x, caption, digits = 3) {
  kableExtra::kbl(x, caption = caption, digits = digits, row.names = FALSE) |>
    kableExtra::kable_styling(
      bootstrap_options = c("striped", "hover", "condensed"),
      full_width = FALSE
    )
}

clean_pathway_name <- function(x) {
  x |>
    stringr::str_remove("^HALLMARK_") |>
    stringr::str_remove("^GOBP_") |>
    stringr::str_remove("^GOCC_") |>
    stringr::str_remove("^GOMF_") |>
    stringr::str_remove("^KEGG_") |>
    stringr::str_replace_all("_", " ") |>
    stringr::str_to_sentence()
}
```

# Metadata

```{r metadata}
# --------------------------------------------------------------------------
# SAMPLE SELECTION
# Only the 11 samples from the three dietary groups are used. All other
# samples present in the count matrix are silently ignored.
# condition_code uses underscores (required by DESeq2 formula syntax);
# condition_label uses the human-readable "ROO+omega3" form for plots.
# --------------------------------------------------------------------------
selected_samples <- c(
  "sample_09", "sample_11", "sample_12",
  "sample_05", "sample_06", "sample_07", "sample_08",
  "sample_17", "sample_18", "sample_19", "sample_20"
)

sample_meta <- tibble::tibble(
  sample = selected_samples,
  mouse_id = c("9", "11", "12", "5", "6", "7", "8", "17", "18", "19", "20"),
  condition_code = factor(
    c(rep("Control", 3), rep("ROO", 4), rep("ROO_omega3", 4)),
    levels = c("Control", "ROO", "ROO_omega3")
  ),
  condition_label = factor(
    c(rep("Control", 3), rep("ROO", 4), rep("ROO+omega3", 4)),
    levels = c("Control", "ROO", "ROO+omega3")
  ),
  replicate = c(1:3, 1:4, 1:4)
)
sample_meta <- as.data.frame(sample_meta)
rownames(sample_meta) <- sample_meta$sample

# Consistent colour palette used in all QC and expression plots
cond_colors <- c(
  Control = "#4D4D4D",
  ROO = "#1B7837",
  `ROO+omega3` = "#762A83"
)

# --------------------------------------------------------------------------
# CONTRAST DEFINITIONS
# Three pairwise contrasts are evaluated. The numerator/denominator convention
# follows DESeq2: log2FC = log2(numerator / denominator).
# All downstream heatmaps are contrast-level: each column represents one
# contrast (not an individual sample). This design choice makes cross-contrast
# patterns immediately visible and avoids inflating figure count.
# --------------------------------------------------------------------------
contrast_table <- tibble::tribble(
  ~label, ~numerator, ~denominator, ~question,
  "ROO vs Control", "ROO", "Control", "Effect of ROO diet versus control diet in mouse heart.",
  "ROO_omega3 vs ROO", "ROO_omega3", "ROO", "Specific effect of omega-3 supplementation on the ROO dietary background.",
  "ROO_omega3 vs Control", "ROO_omega3", "Control", "Effect of ROO plus omega-3 diet versus control diet in mouse heart."
)

contrast_order <- contrast_table$label
contrast_display_names <- c(
  "ROO vs Control" = "ROO vs Control",
  "ROO_omega3 vs ROO" = "ROO+w3 vs ROO",
  "ROO_omega3 vs Control" = "ROO+w3 vs Control"
)

# Shared up/down/neutral palette for volcano plots, chord plots and heatmap annotations
up_down_colors <- c(
  Up = "#B2182B",
  Down = "#2166AC",
  Neutral = "grey70"
)

make_kable(sample_meta, "Mouse heart diet metadata")
make_kable(contrast_table, "Contrasts - mouse heart diet analysis")
```

# VERTIENTE 1. RNA-seq QC

## Load count matrix and select heart diet samples

```{r load-counts}
# --------------------------------------------------------------------------
# COUNT MATRIX LOADING AND PREPROCESSING
# The first column is coerced to "gene_id" regardless of its original name.
# Gene IDs may be either plain Ensembl IDs (ENSMUSG...) or composite strings
# of the form "ENSMUSG...|GeneSymbol"; both formats are handled.
# Version suffixes (.N) are stripped from Ensembl IDs before annotation.
# Counts for duplicate Ensembl IDs (e.g. from multi-mapping) are summed.
# The final count_matrix is an integer matrix with genes as rows and samples
# as columns, aligned to the row order of sample_meta.
# --------------------------------------------------------------------------
if (!file.exists(COUNT_FILE)) stop("Count file not found: ", COUNT_FILE)

counts_tbl <- readr::read_csv(COUNT_FILE, show_col_types = FALSE)

if (!"gene_id" %in% colnames(counts_tbl) && "Geneid" %in% colnames(counts_tbl)) {
  counts_tbl <- counts_tbl |> dplyr::rename(gene_id = Geneid)
}
if (!"gene_id" %in% colnames(counts_tbl)) {
  colnames(counts_tbl)[1] <- "gene_id"
}

missing_samples <- setdiff(selected_samples, colnames(counts_tbl))
if (length(missing_samples) > 0) {
  stop("Missing selected samples in count matrix: ", paste(missing_samples, collapse = ", "))
}

counts_tbl <- counts_tbl |>
  dplyr::select(gene_id, dplyr::all_of(selected_samples)) |>
  dplyr::mutate(
    ensembl_id = stringr::str_replace(gene_id, "\\.[0-9]+$", ""),
    ensembl_id = stringr::str_replace(ensembl_id, "\\|.*$", ""),
    input_symbol = dplyr::if_else(
      stringr::str_detect(gene_id, "\\|"),
      stringr::str_replace(gene_id, "^.*\\|", ""),
      NA_character_
    ),
    input_symbol = dplyr::na_if(input_symbol, "")
  )

symbol_map_from_file <- counts_tbl |>
  dplyr::select(ensembl_id, input_symbol) |>
  dplyr::filter(!is.na(input_symbol), input_symbol != "") |>
  dplyr::distinct(ensembl_id, .keep_all = TRUE)

counts_aggregated <- counts_tbl |>
  dplyr::select(ensembl_id, dplyr::all_of(selected_samples)) |>
  dplyr::group_by(ensembl_id) |>
  dplyr::summarise(dplyr::across(dplyr::all_of(selected_samples), ~ sum(as.numeric(.x), na.rm = TRUE)), .groups = "drop")

count_matrix <- counts_aggregated |>
  dplyr::select(dplyr::all_of(selected_samples)) |>
  as.matrix()
rownames(count_matrix) <- counts_aggregated$ensembl_id
storage.mode(count_matrix) <- "integer"

stopifnot(all(colnames(count_matrix) == rownames(sample_meta)))
stopifnot(!anyNA(count_matrix), all(count_matrix >= 0))

cat("Mouse heart diet matrix:", nrow(count_matrix), "genes x", ncol(count_matrix), "samples\n")
cat("Genes with zero counts across selected samples:", sum(rowSums(count_matrix) == 0), "\n")
make_kable(head(as.data.frame(count_matrix[, seq_len(min(6, ncol(count_matrix)))])), "First rows after sample selection")
```

## Library size, detected genes and raw count distributions

```{r qc-basic, fig.width=10, fig.height=5}
qc_metrics <- tibble::tibble(
  sample = colnames(count_matrix),
  library_size = colSums(count_matrix),
  library_size_millions = library_size / 1e6,
  detected_genes = colSums(count_matrix > 0)
) |>
  dplyr::left_join(sample_meta |> dplyr::select(sample, condition_label), by = "sample")

readr::write_csv(qc_metrics, file.path(DIRS$qc, "sample_qc_metrics_mouse_heart_diet.csv"))

p_lib <- ggplot(qc_metrics, aes(sample, library_size_millions, fill = condition_label)) +
  geom_col(color = "white", linewidth = 0.3) +
  geom_hline(yintercept = median(qc_metrics$library_size_millions), linetype = "dashed", color = "grey35") +
  scale_fill_manual(values = cond_colors, name = "Condition") +
  scale_y_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.08))) +
  labs(title = "Library size - mouse heart diet samples", x = NULL, y = "Total counts (millions)") +
  heart_theme() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p_detected <- ggplot(qc_metrics, aes(sample, detected_genes, fill = condition_label)) +
  geom_col(color = "white", linewidth = 0.3) +
  scale_fill_manual(values = cond_colors, name = "Condition") +
  scale_y_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.08))) +
  labs(title = "Detected genes - mouse heart diet samples", x = NULL, y = "Genes with count > 0") +
  heart_theme() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

print(p_lib)
print(p_detected)
save_plot(p_lib, "QC_library_size_mouse_heart_diet", width = 10, height = 5, dir = DIRS$qc)
save_plot(p_detected, "QC_detected_genes_mouse_heart_diet", width = 10, height = 5, dir = DIRS$qc)

log_counts <- log2(count_matrix + 1)
df_box <- as.data.frame(log_counts) |>
  tibble::rownames_to_column("ensembl_id") |>
  tidyr::pivot_longer(-ensembl_id, names_to = "sample", values_to = "log2_count") |>
  dplyr::left_join(sample_meta |> dplyr::select(sample, condition_label), by = "sample")

p_box <- ggplot(df_box, aes(sample, log2_count, fill = condition_label)) +
  geom_boxplot(outlier.size = 0.2, outlier.alpha = 0.25, linewidth = 0.35) +
  scale_fill_manual(values = cond_colors, name = "Condition") +
  labs(title = "Raw count distribution - mouse heart diet samples", x = NULL, y = expression(log[2](count + 1))) +
  heart_theme() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
print(p_box)
save_plot(p_box, "QC_raw_distribution_mouse_heart_diet", width = 10, height = 5, dir = DIRS$qc)

make_kable(qc_metrics, "QC metrics - mouse heart diet samples")
```

## Filtering, DESeq2 normalisation and PCA

```{r deseq-qc, fig.width=8, fig.height=6}
# --------------------------------------------------------------------------
# LOW-EXPRESSION FILTERING AND DESeq2 NORMALISATION
# Genes with fewer than MIN_COUNT reads in at least MIN_SAMPLES samples are
# removed before fitting the DESeq2 model. This conservative filter reduces
# multiple-testing burden without discarding biologically relevant lowly
# expressed genes.
#
# DESeq2 is run with a simple ~ condition_code design. Normalised expression
# values for QC and visualisation are obtained via variance-stabilising
# transformation (VST, blind = TRUE), which is appropriate for exploratory
# analyses because it does not use the design matrix.
#
# PCA is computed on both the raw log2(count+1) matrix (before normalisation)
# and the VST-normalised matrix to show the effect of normalisation on
# sample clustering.
# --------------------------------------------------------------------------
keep_gene <- rowSums(count_matrix >= MIN_COUNT) >= MIN_SAMPLES
count_matrix_filtered <- count_matrix[keep_gene, , drop = FALSE]

filter_summary <- tibble::tibble(
  metric = c("Genes before filtering", "Genes retained", "Genes removed"),
  value = c(nrow(count_matrix), nrow(count_matrix_filtered), nrow(count_matrix) - nrow(count_matrix_filtered))
)
readr::write_csv(filter_summary, file.path(DIRS$qc, "filtering_summary_mouse_heart_diet.csv"))
make_kable(filter_summary, "Filtering summary - mouse heart diet samples")

dds <- DESeq2::DESeqDataSetFromMatrix(
  countData = count_matrix_filtered,
  colData = sample_meta,
  design = ~ condition_code
)
dds$condition_code <- stats::relevel(dds$condition_code, ref = "Control")
dds <- DESeq2::DESeq(dds)

size_factor_table <- tibble::tibble(
  sample = names(DESeq2::sizeFactors(dds)),
  size_factor = as.numeric(DESeq2::sizeFactors(dds))
) |>
  dplyr::left_join(sample_meta |> dplyr::select(sample, condition_label), by = "sample")
readr::write_csv(size_factor_table, file.path(DIRS$qc, "size_factors_mouse_heart_diet.csv"))

vst_data <- DESeq2::vst(dds, blind = TRUE)
vst_mat <- SummarizedExperiment::assay(vst_data)

run_pca <- function(mat, title) {
  pca <- prcomp(t(mat), center = TRUE, scale. = TRUE)
  pct <- round(100 * pca$sdev^2 / sum(pca$sdev^2), 1)
  df <- as.data.frame(pca$x[, 1:2]) |>
    tibble::rownames_to_column("sample") |>
    dplyr::left_join(sample_meta |> dplyr::select(sample, condition_label, mouse_id), by = "sample")
  ggplot(df, aes(PC1, PC2, color = condition_label, label = mouse_id)) +
    geom_point(size = 3.2) +
    ggrepel::geom_text_repel(size = 3, max.overlaps = Inf, seed = 42) +
    scale_color_manual(values = cond_colors, name = "Condition") +
    labs(
      title = title,
      x = paste0("PC1: ", pct[1], "% variance"),
      y = paste0("PC2: ", pct[2], "% variance")
    ) +
    coord_fixed() +
    heart_theme()
}

p_pca_raw <- run_pca(log_counts[rownames(count_matrix_filtered), ], "PCA before normalisation - mouse heart diet samples")
p_pca_vst <- run_pca(vst_mat, "PCA after VST normalisation - mouse heart diet samples")
print(p_pca_raw)
print(p_pca_vst)
save_plot(p_pca_raw, "QC_PCA_raw_mouse_heart_diet", width = 8, height = 6, dir = DIRS$qc)
save_plot(p_pca_vst, "QC_PCA_VST_mouse_heart_diet", width = 8, height = 6, dir = DIRS$qc)
```

## Sample distances and outlier metrics

```{r sample-distance, fig.width=8, fig.height=7}
# --------------------------------------------------------------------------
# SAMPLE SIMILARITY AND OUTLIER DETECTION
# Euclidean distances and Pearson correlations are computed on the VST matrix.
# A sample is flagged as a potential outlier if its median distance to all
# other samples exceeds 2 SD above the group mean, or if its mean correlation
# falls more than 2 SD below the mean (distance_z > 2 | correlation_z < -2).
#
# The QC heatmap below is a contrast-level summary: instead of plotting a
# sample × sample matrix, columns represent contrasts and values are the
# difference between group centroids in VST space
# (mean VST numerator - mean VST denominator). This makes the QC heatmap
# directly comparable to the downstream expression heatmaps.
# --------------------------------------------------------------------------
sample_dist <- dist(t(vst_mat), method = "euclidean")
sample_dist_mat <- as.matrix(sample_dist)
sample_cor <- cor(vst_mat, method = "pearson")

outlier_metrics <- tibble::tibble(
  sample = colnames(vst_mat),
  median_distance_to_others = apply(sample_dist_mat, 1, function(x) median(x[x > 0])),
  mean_correlation_to_others = apply(sample_cor, 1, function(x) mean(x[x < 0.999999]))
) |>
  dplyr::mutate(
    distance_z = as.numeric(scale(median_distance_to_others)),
    correlation_z = as.numeric(scale(mean_correlation_to_others)),
    potential_outlier = distance_z > 2 | correlation_z < -2
  ) |>
  dplyr::left_join(sample_meta |> dplyr::select(sample, condition_label, mouse_id), by = "sample")
readr::write_csv(outlier_metrics, file.path(DIRS$qc, "sample_outlier_metrics_mouse_heart_diet.csv"))
make_kable(outlier_metrics, "Outlier metrics - mouse heart diet samples")

# Instead of sample-by-sample heatmaps, this QC heatmap summarises the same VST data
# at contrast level. Values are differences between group centroids in VST space:
# mean(VST numerator) - mean(VST denominator). Therefore, every column is a contrast.
make_vst_centroid_contrast_matrix <- function(expr_mat, genes_use = rownames(expr_mat)) {
  genes_use <- intersect(genes_use, rownames(expr_mat))
  out <- sapply(seq_len(nrow(contrast_table)), function(i) {
    num <- contrast_table$numerator[i]
    den <- contrast_table$denominator[i]
    num_samples <- rownames(sample_meta)[sample_meta$condition_code == num]
    den_samples <- rownames(sample_meta)[sample_meta$condition_code == den]
    rowMeans(expr_mat[genes_use, num_samples, drop = FALSE]) -
      rowMeans(expr_mat[genes_use, den_samples, drop = FALSE])
  })
  colnames(out) <- contrast_display_names[contrast_table$label]
  rownames(out) <- genes_use
  out
}

top_var_genes_qc <- names(sort(apply(vst_mat, 1, var), decreasing = TRUE))[seq_len(min(150, nrow(vst_mat)))]
qc_contrast_mat <- make_vst_centroid_contrast_matrix(vst_mat, top_var_genes_qc)
qc_contrast_mat_z <- t(scale(t(qc_contrast_mat)))
qc_contrast_mat_z[is.nan(qc_contrast_mat_z)] <- 0

# Row names are Ensembl IDs at this stage (gene_annotation is built later in
# Vertiente 2). They are not displayed because show_row_names = FALSE, so no
# translation to symbols is needed here.

ht_qc <- ComplexHeatmap::Heatmap(
  qc_contrast_mat_z,
  name = "Z-score",
  col = circlize::colorRamp2(c(-2, 0, 2), c("#2166AC", "white", "#B2182B")),
  cluster_rows = TRUE,
  cluster_columns = FALSE,
  show_row_names = FALSE,
  column_names_rot = 45,
  column_title = "QC contrast-level VST centroid differences",
  heatmap_legend_param = list(title = "Scaled\ncentroid\ndifference")
)

fn_qc_svg <- file.path(DIRS$qc, "QC_heatmap_VST_centroid_differences_by_contrast.svg")
fn_qc_png <- file.path(DIRS$qc, "QC_heatmap_VST_centroid_differences_by_contrast.png")
grDevices::svg(fn_qc_svg, width = 7, height = 8)
ComplexHeatmap::draw(ht_qc)
grDevices::dev.off()
png(fn_qc_png, width = 7, height = 8, units = "in", res = 300)
ComplexHeatmap::draw(ht_qc)
dev.off()
ComplexHeatmap::draw(ht_qc)
```

# VERTIENTE 2. Differential expression

## Mouse annotation and contrast extraction

```{r de-results}
# --------------------------------------------------------------------------
# GENE ANNOTATION (org.Mm.eg.db)
# Ensembl IDs are mapped to gene symbols, full gene names, biotype and
# Entrez IDs using the org.Mm.eg.db annotation package. When the database
# returns no symbol (e.g. novel or unannotated transcripts), the script
# falls back to the symbol embedded in the count matrix gene_id column;
# if that is also absent, the Ensembl ID itself is used as the display name.
#
# CONTRAST EXTRACTION (DESeq2::results)
# Each contrast is extracted independently with pAdjustMethod = "BH".
# Genes are classified as Up / Down / NS using the global PADJ_CUTOFF and
# LFC_CUTOFF thresholds. The direction column (Up/Down/Neutral) uses only
# the sign of log2FC and is used in chord plots regardless of significance.
#
# HELPER MATRICES (build_lfc_matrix / build_padj_matrix)
# Both functions return a gene × contrast matrix ready to pass to
# ComplexHeatmap. Log2FC values are capped at ±cap to prevent extreme
# values from dominating the colour scale. Significance asterisks
# (* p<0.05, ** p<0.01, *** p<0.001) are overlaid on each heatmap cell
# by make_sig_labels().
# --------------------------------------------------------------------------
annotate_ensembl_mouse <- function(ensembl_ids, symbol_map_from_file = NULL) {
  available_cols <- AnnotationDbi::columns(org.Mm.eg.db)
  wanted_cols <- intersect(c("SYMBOL", "GENENAME", "GENETYPE", "ENTREZID"), available_cols)

  annot <- AnnotationDbi::select(
    org.Mm.eg.db,
    keys = unique(ensembl_ids),
    keytype = "ENSEMBL",
    columns = wanted_cols
  )
  annot <- annot[!duplicated(annot$ENSEMBL), , drop = FALSE]
  annot <- annot |>
    dplyr::rename(ensembl_id = ENSEMBL)

  if (!"SYMBOL" %in% colnames(annot)) annot$SYMBOL <- NA_character_
  if (!"GENENAME" %in% colnames(annot)) annot$GENENAME <- NA_character_
  if (!"GENETYPE" %in% colnames(annot)) annot$GENETYPE <- NA_character_
  if (!"ENTREZID" %in% colnames(annot)) annot$ENTREZID <- NA_character_

  annot <- annot |>
    dplyr::rename(
      symbol_orgdb = SYMBOL,
      description = GENENAME,
      gene_biotype = GENETYPE,
      entrez_id = ENTREZID
    )

  base <- tibble::tibble(ensembl_id = unique(ensembl_ids))
  out <- base |>
    dplyr::left_join(annot, by = "ensembl_id")

  if (!is.null(symbol_map_from_file)) {
    out <- out |>
      dplyr::left_join(symbol_map_from_file, by = "ensembl_id")
  } else {
    out$input_symbol <- NA_character_
  }

  out |>
    dplyr::mutate(
      symbol = dplyr::coalesce(symbol_orgdb, input_symbol, ensembl_id),
      symbol = dplyr::if_else(is.na(symbol) | symbol == "", ensembl_id, symbol),
      symbol_upper = toupper(symbol)
    ) |>
    dplyr::select(ensembl_id, symbol, symbol_upper, description, gene_biotype, entrez_id, input_symbol, symbol_orgdb)
}

gene_annotation <- annotate_ensembl_mouse(rownames(dds), symbol_map_from_file)
readr::write_csv(gene_annotation, file.path(DIRS$tables, "gene_annotation_mouse_heart_diet.csv"))
readr::write_csv(contrast_table, file.path(DIRS$de, "contrast_definitions_mouse_heart_diet.csv"))

extract_contrast <- function(numerator, denominator, label) {
  res <- DESeq2::results(
    dds,
    contrast = c("condition_code", numerator, denominator),
    pAdjustMethod = "BH",
    alpha = PADJ_CUTOFF
  )

  as.data.frame(res) |>
    tibble::rownames_to_column("ensembl_id") |>
    dplyr::left_join(gene_annotation, by = "ensembl_id") |>
    dplyr::mutate(
      symbol = dplyr::if_else(is.na(symbol) | symbol == "", ensembl_id, symbol),
      contrast = label,
      status = dplyr::case_when(
        !is.na(padj) & padj < PADJ_CUTOFF & log2FoldChange >= LFC_CUTOFF ~ "Up",
        !is.na(padj) & padj < PADJ_CUTOFF & log2FoldChange <= -LFC_CUTOFF ~ "Down",
        TRUE ~ "NS"
      ),
      direction = dplyr::case_when(
        !is.na(log2FoldChange) & log2FoldChange > 0 ~ "Up",
        !is.na(log2FoldChange) & log2FoldChange < 0 ~ "Down",
        TRUE ~ "Neutral"
      )
    ) |>
    dplyr::arrange(padj)
}

contrast_results <- setNames(vector("list", nrow(contrast_table)), contrast_table$label)
for (i in seq_len(nrow(contrast_table))) {
  ct <- contrast_table[i, ]
  contrast_results[[ct$label]] <- extract_contrast(ct$numerator, ct$denominator, ct$label)
}

for (nm in names(contrast_results)) {
  readr::write_csv(
    contrast_results[[nm]],
    file.path(DIRS$de, paste0("DESeq2_all_", safe_name(nm), ".csv"))
  )
  readr::write_csv(
    contrast_results[[nm]] |> dplyr::filter(status != "NS"),
    file.path(DIRS$de, paste0("DESeq2_DEGs_", safe_name(nm), ".csv"))
  )
}

deg_summary <- dplyr::bind_rows(lapply(names(contrast_results), function(nm) {
  tab <- contrast_results[[nm]]
  tibble::tibble(
    contrast = nm,
    upregulated = sum(tab$status == "Up", na.rm = TRUE),
    downregulated = sum(tab$status == "Down", na.rm = TRUE),
    total_degs = sum(tab$status != "NS", na.rm = TRUE)
  )
}))
readr::write_csv(deg_summary, file.path(DIRS$de, "DEG_summary_mouse_heart_diet.csv"))
make_kable(deg_summary, "DEG summary - mouse heart diet analysis")

# Helper matrices for all contrast-level gene heatmaps.
# Rows = genes; columns = contrasts; values = DESeq2 log2 fold-change.
build_lfc_matrix <- function(genes_use, row_labels = NULL, cap = 3) {
  genes_use <- unique(genes_use[!is.na(genes_use)])

  lfc_df <- dplyr::bind_rows(lapply(contrast_order, function(nm) {
    contrast_results[[nm]] |>
      dplyr::filter(ensembl_id %in% genes_use) |>
      dplyr::select(ensembl_id, symbol, contrast, log2FoldChange, padj, status, direction)
  })) |>
    dplyr::mutate(
      contrast = factor(contrast, levels = contrast_order),
      contrast_label = contrast_display_names[as.character(contrast)]
    )

  mat <- lfc_df |>
    dplyr::select(ensembl_id, contrast_label, log2FoldChange) |>
    tidyr::pivot_wider(names_from = contrast_label, values_from = log2FoldChange) |>
    tibble::column_to_rownames("ensembl_id") |>
    as.matrix()

  mat <- mat[intersect(genes_use, rownames(mat)), unname(contrast_display_names[contrast_order]), drop = FALSE]
  mat[is.infinite(mat)] <- NA_real_
  mat_plot <- pmax(pmin(mat, cap), -cap)

  if (is.null(row_labels)) {
    row_labels <- gene_annotation$symbol[match(rownames(mat_plot), gene_annotation$ensembl_id)]
  }
  row_labels <- dplyr::if_else(is.na(row_labels) | row_labels == "", rownames(mat_plot), row_labels)
  rownames(mat_plot) <- make.unique(row_labels)
  mat_plot
}

build_padj_matrix <- function(genes_use, row_labels = NULL) {
  genes_use <- unique(genes_use[!is.na(genes_use)])

  padj_df <- dplyr::bind_rows(lapply(contrast_order, function(nm) {
    contrast_results[[nm]] |>
      dplyr::filter(ensembl_id %in% genes_use) |>
      dplyr::select(ensembl_id, contrast, padj)
  })) |>
    dplyr::mutate(
      contrast = factor(contrast, levels = contrast_order),
      contrast_label = contrast_display_names[as.character(contrast)]
    )

  mat <- padj_df |>
    dplyr::select(ensembl_id, contrast_label, padj) |>
    tidyr::pivot_wider(names_from = contrast_label, values_from = padj) |>
    tibble::column_to_rownames("ensembl_id") |>
    as.matrix()

  mat <- mat[intersect(genes_use, rownames(mat)), unname(contrast_display_names[contrast_order]), drop = FALSE]

  if (is.null(row_labels)) {
    row_labels <- gene_annotation$symbol[match(rownames(mat), gene_annotation$ensembl_id)]
  }
  row_labels <- dplyr::if_else(is.na(row_labels) | row_labels == "", rownames(mat), row_labels)
  rownames(mat) <- make.unique(row_labels)
  mat
}

make_sig_labels <- function(padj_mat) {
  sig_mat <- matrix("", nrow = nrow(padj_mat), ncol = ncol(padj_mat), dimnames = dimnames(padj_mat))
  sig_mat[!is.na(padj_mat) & padj_mat < 0.05] <- "*"
  sig_mat[!is.na(padj_mat) & padj_mat < 0.01] <- "**"
  sig_mat[!is.na(padj_mat) & padj_mat < 0.001] <- "***"
  sig_mat
}
```

## Volcano and MA plots

```{r volcano-ma, fig.width=8, fig.height=6}
make_volcano <- function(res_df, title) {
  plot_df <- res_df |>
    dplyr::mutate(
      neg_log10_padj = -log10(padj),
      status = factor(status, levels = c("Down", "NS", "Up")),
      label_gene = !stringr::str_detect(symbol, "^ENSMUSG")
    )

  max_finite <- suppressWarnings(max(plot_df$neg_log10_padj[is.finite(plot_df$neg_log10_padj)], na.rm = TRUE))
  if (!is.finite(max_finite)) max_finite <- 0
  plot_df <- plot_df |>
    dplyr::mutate(neg_log10_padj = ifelse(is.infinite(neg_log10_padj), max_finite + 1, neg_log10_padj))

  labels <- dplyr::bind_rows(
    plot_df |> dplyr::filter(status == "Up", label_gene) |> dplyr::arrange(padj) |> dplyr::slice_head(n = TOP_LABEL_GENES),
    plot_df |> dplyr::filter(status == "Down", label_gene) |> dplyr::arrange(padj) |> dplyr::slice_head(n = TOP_LABEL_GENES)
  )

  pal <- c(Down = "#2166AC", NS = "grey78", Up = "#B2182B")
  ggplot(plot_df, aes(log2FoldChange, neg_log10_padj)) +
    geom_point(aes(color = status), size = 1.2, alpha = 0.75, na.rm = TRUE) +
    geom_vline(xintercept = c(-LFC_CUTOFF, LFC_CUTOFF), linetype = "dashed", color = "grey45") +
    geom_hline(yintercept = -log10(PADJ_CUTOFF), linetype = "dashed", color = "grey45") +
    ggrepel::geom_text_repel(
      data = labels,
      aes(label = symbol, color = status),
      size = 2.8,
      max.overlaps = Inf,
      seed = 42,
      fontface = "italic",
      show.legend = FALSE
    ) +
    scale_color_manual(values = pal, name = NULL) +
    labs(
      title = title,
      subtitle = paste0("|log2FC| >= ", LFC_CUTOFF, "; adjusted p < ", PADJ_CUTOFF),
      x = expression(log[2]~fold~change),
      y = expression(-log[10]~adjusted~italic(p))
    ) +
    heart_theme()
}

make_ma <- function(res_df, title) {
  pal <- c(Down = "#2166AC", NS = "grey78", Up = "#B2182B")
  ggplot(res_df, aes(log10(baseMean + 1), log2FoldChange, color = status)) +
    geom_point(size = 0.8, alpha = 0.55, na.rm = TRUE) +
    geom_hline(yintercept = 0, color = "black", linewidth = 0.4) +
    geom_hline(yintercept = c(-LFC_CUTOFF, LFC_CUTOFF), linetype = "dashed", color = "grey45") +
    scale_color_manual(values = pal, name = NULL) +
    labs(
      title = title,
      x = expression(log[10]~(mean~normalized~counts + 1)),
      y = expression(log[2]~fold~change)
    ) +
    heart_theme()
}

for (nm in names(contrast_results)) {
  p_v <- make_volcano(contrast_results[[nm]], nm)
  p_m <- make_ma(contrast_results[[nm]], nm)
  print(p_v)
  print(p_m)
  save_plot(p_v, paste0("Volcano_", safe_name(nm)), width = 8, height = 6, dir = DIRS$de)
  save_plot(p_m, paste0("MA_", safe_name(nm)), width = 8, height = 5, dir = DIRS$de)
}
```

## DEG heatmaps

```{r deg-heatmaps, fig.width=10, fig.height=10}
# Union of the strongest DEGs from each contrast.
# This is a contrast-level heatmap: columns are ROO vs Control,
# ROO+w3 vs ROO and ROO+w3 vs Control. Values are log2 fold-change.
selected_deg_genes <- dplyr::bind_rows(lapply(contrast_order, function(nm) {
  contrast_results[[nm]] |>
    dplyr::filter(status != "NS", !is.na(padj)) |>
    dplyr::arrange(padj, dplyr::desc(abs(log2FoldChange))) |>
    dplyr::slice_head(n = TOP_HEATMAP_GENES) |>
    dplyr::mutate(selection_contrast = nm)
})) |>
  dplyr::group_by(ensembl_id, symbol) |>
  dplyr::summarise(
    best_padj = min(padj, na.rm = TRUE),
    strongest_abs_lfc = max(abs(log2FoldChange), na.rm = TRUE),
    main_selection_contrast = selection_contrast[which.min(padj)][1],
    .groups = "drop"
  ) |>
  dplyr::arrange(best_padj, dplyr::desc(strongest_abs_lfc)) |>
  dplyr::slice_head(n = max(TOP_HEATMAP_GENES, 75))

if (nrow(selected_deg_genes) >= 2) {
  genes_use <- selected_deg_genes$ensembl_id
  labels_use <- selected_deg_genes$symbol

  lfc_mat <- build_lfc_matrix(genes_use, row_labels = labels_use, cap = 3)
  padj_mat <- build_padj_matrix(genes_use, row_labels = labels_use)
  sig_mat <- make_sig_labels(padj_mat)

  # Rows are clustered by their log2FC pattern across contrasts.
  # We do not split by samples because this heatmap is intentionally contrast-level.
  row_split <- NULL

  ht <- ComplexHeatmap::Heatmap(
    lfc_mat,
    name = "log2FC",
    col = circlize::colorRamp2(c(-3, 0, 3), c("#2166AC", "white", "#B2182B")),
    na_col = "grey92",
    cluster_rows = TRUE,
    cluster_columns = FALSE,
    row_split = row_split,
    row_names_gp = grid::gpar(fontsize = 7, fontface = "italic"),
    column_names_rot = 45,
    column_title = "Top DEGs by contrast - mouse heart diet",
    rect_gp = grid::gpar(col = "white", lwd = 0.4),
    cell_fun = function(j, i, x, y, width, height, fill) {
      if (sig_mat[i, j] != "") {
        grid::grid.text(sig_mat[i, j], x = x, y = y, gp = grid::gpar(fontsize = 7, fontface = "bold"))
      }
    },
    heatmap_legend_param = list(title = "log2FC")
  )

  fn_svg <- file.path(DIRS$de, "Heatmap_top_DEGs_by_contrast_mouse_heart_diet.svg")
  fn_png <- file.path(DIRS$de, "Heatmap_top_DEGs_by_contrast_mouse_heart_diet.png")
  grDevices::svg(fn_svg, width = 8, height = max(7, 0.18 * nrow(lfc_mat) + 2))
  ComplexHeatmap::draw(ht)
  grDevices::dev.off()
  png(fn_png, width = 8, height = max(7, 0.18 * nrow(lfc_mat) + 2), units = "in", res = 300)
  ComplexHeatmap::draw(ht)
  dev.off()
  ComplexHeatmap::draw(ht)

  readr::write_csv(
    tibble::rownames_to_column(as.data.frame(lfc_mat), "gene_symbol"),
    file.path(DIRS$de, "Heatmap_top_DEGs_by_contrast_log2FC_matrix.csv")
  )
} else {
  message("No significant DEGs available for a contrast-level DEG heatmap.")
}
```

# VERTIENTE 3. Targeted cardiac biology

## Curated cardiac gene panels

```{r targeted-panels}
# --------------------------------------------------------------------------
# CURATED CARDIAC GENE PANELS
# Ten biological processes relevant to cardiac physiology and cardiometabolic
# disease are assessed using manually curated gene lists. The panels cover:
# contractility/calcium handling, hypertrophic stress response, fatty acid
# oxidation and lipid handling, mitochondrial OXPHOS, glucose/insulin
# signalling, inflammation/immune infiltration, fibrosis/ECM remodelling,
# oxidative stress/antioxidant response, ER stress/autophagy/proteostasis,
# and endothelial/vascular response.
#
# Gene symbols are matched case-insensitively against the annotation table;
# symbols absent from the filtered count matrix are skipped silently.
# A separate list (interesting_genes) drives the individual boxplots in
# the next section.
# --------------------------------------------------------------------------
heart_gene_sets <- list(
  Cardiac_contractility_and_calcium = c(
    "Myh6", "Myh7", "Tnnt2", "Tnni3", "Tpm1", "Actn2", "Mybpc3", "Atp2a2",
    "Ryr2", "Pln", "Cacna1c", "Casq2", "Slc8a1"
  ),
  Hypertrophy_and_stress_response = c(
    "Nppa", "Nppb", "Myh7", "Acta1", "Ankrd1", "Fhl1", "Xirp2", "Fos", "Jun", "Egr1"
  ),
  Fatty_acid_oxidation_and_lipid_handling = c(
    "Ppara", "Ppard", "Ppargc1a", "Cpt1b", "Acadvl", "Acadm", "Hadha", "Hadh",
    "Cd36", "Lpl", "Fabp3", "Plin5", "Acox1", "Scd1", "Ldlr", "Hmgcr"
  ),
  Mitochondrial_OXPHOS = c(
    "Ndufa9", "Ndufb8", "Sdha", "Uqcrc2", "Cox4i1", "Cox5a", "Atp5f1a", "Atp5f1b",
    "Tfam", "Tomm20", "Cycs", "Ppargc1a"
  ),
  Glucose_metabolism_and_insulin_signalling = c(
    "Slc2a1", "Slc2a4", "Hk2", "Pfkfb3", "Pgam1", "Pdk4", "Ldha", "Insr", "Irs1", "Akt1"
  ),
  Inflammation_and_immune_cells = c(
    "Il1b", "Il6", "Tnf", "Ccl2", "Cxcl10", "Nfkb1", "Nfkbia", "Nlrp3", "Ptprc",
    "Adgre1", "Itgam", "Cd68", "Lgals3", "Ccr2"
  ),
  Fibrosis_and_ECM_remodelling = c(
    "Col1a1", "Col1a2", "Col3a1", "Col6a1", "Fn1", "Tgfb1", "Postn", "Acta2",
    "Mmp2", "Mmp9", "Timp1", "Timp2", "Lox"
  ),
  Oxidative_stress_and_antioxidant_response = c(
    "Nfe2l2", "Keap1", "Hmox1", "Nqo1", "Sod1", "Sod2", "Cat", "Gpx1", "Gpx4", "Prdx1", "Txnrd1", "Nos2"
  ),
  ER_stress_autophagy_and_proteostasis = c(
    "Ddit3", "Atf4", "Atf6", "Hspa5", "Xbp1", "Eif2ak3", "Map1lc3b", "Sqstm1", "Becn1", "Atg5"
  ),
  Endothelial_and_vascular_response = c(
    "Pecam1", "Vwf", "Kdr", "Flt1", "Klf2", "Nos3", "Vegfa", "Tek", "Cdh5", "Icam1", "Vcam1"
  )
)

interesting_genes <- unique(c(
  "Nppa", "Nppb", "Myh6", "Myh7", "Tnnt2", "Atp2a2", "Ryr2", "Pln",
  "Ppara", "Ppargc1a", "Cpt1b", "Acadvl", "Cd36", "Lpl", "Fabp3",
  "Slc2a4", "Pdk4", "Col1a1", "Col3a1", "Tgfb1", "Postn", "Acta2",
  "Il6", "Tnf", "Il1b", "Ccl2", "Cxcl10", "Cd68", "Lgals3",
  "Nfe2l2", "Hmox1", "Nqo1", "Sod2", "Gpx4", "Nos2", "Pecam1", "Vwf", "Nos3"
))

symbol_to_ensembl <- gene_annotation |>
  dplyr::filter(!is.na(symbol), symbol != "") |>
  dplyr::select(symbol, symbol_upper, ensembl_id) |>
  dplyr::distinct(symbol_upper, .keep_all = TRUE)

map_gene_symbols <- function(gene_symbols) {
  tibble::tibble(target_symbol = gene_symbols, symbol_upper = toupper(gene_symbols)) |>
    dplyr::left_join(symbol_to_ensembl, by = "symbol_upper") |>
    dplyr::mutate(present = !is.na(ensembl_id) & ensembl_id %in% rownames(vst_mat))
}

target_presence <- dplyr::bind_rows(lapply(names(heart_gene_sets), function(set_name) {
  map_gene_symbols(heart_gene_sets[[set_name]]) |>
    dplyr::mutate(category = set_name, .before = 1)
}))

readr::write_csv(target_presence, file.path(DIRS$targeted, "target_gene_presence_mouse_heart_diet.csv"))
make_kable(target_presence, "Targeted cardiac gene presence")
```

## Targeted cardiac-process heatmaps

```{r targeted-heatmaps, fig.width=12, fig.height=10}
make_process_heatmap <- function(process_name, gene_symbols) {
  mapped <- map_gene_symbols(gene_symbols) |>
    dplyr::filter(present) |>
    dplyr::distinct(ensembl_id, .keep_all = TRUE)

  if (nrow(mapped) < 2) {
    message("Fewer than 2 genes present for process: ", process_name)
    return(invisible(NULL))
  }

  genes_use <- mapped$ensembl_id
  labels_use <- mapped$symbol[match(genes_use, mapped$ensembl_id)]
  lfc_mat <- build_lfc_matrix(genes_use, row_labels = labels_use, cap = 2.5)
  padj_mat <- build_padj_matrix(genes_use, row_labels = labels_use)
  sig_mat <- make_sig_labels(padj_mat)

  ht <- ComplexHeatmap::Heatmap(
    lfc_mat,
    name = "log2FC",
    col = circlize::colorRamp2(c(-2.5, 0, 2.5), c("#2166AC", "white", "#B2182B")),
    na_col = "grey92",
    cluster_rows = TRUE,
    cluster_columns = FALSE,
    column_names_rot = 45,
    row_names_gp = grid::gpar(fontsize = 8, fontface = "italic"),
    column_title = paste0(stringr::str_replace_all(process_name, "_", " "), " - contrast-level log2FC"),
    rect_gp = grid::gpar(col = "white", lwd = 0.4),
    cell_fun = function(j, i, x, y, width, height, fill) {
      if (sig_mat[i, j] != "") {
        grid::grid.text(sig_mat[i, j], x = x, y = y, gp = grid::gpar(fontsize = 7, fontface = "bold"))
      }
    },
    heatmap_legend_param = list(title = "log2FC")
  )

  fn_svg <- file.path(DIRS$targeted, paste0("Heatmap_process_by_contrast_", safe_name(process_name), ".svg"))
  fn_png <- file.path(DIRS$targeted, paste0("Heatmap_process_by_contrast_", safe_name(process_name), ".png"))
  grDevices::svg(fn_svg, width = 8, height = max(5.5, 0.28 * nrow(lfc_mat) + 2))
  ComplexHeatmap::draw(ht)
  grDevices::dev.off()
  png(fn_png, width = 8, height = max(5.5, 0.28 * nrow(lfc_mat) + 2), units = "in", res = 300)
  ComplexHeatmap::draw(ht)
  dev.off()
  ComplexHeatmap::draw(ht)
}

for (process_name in names(heart_gene_sets)) {
  make_process_heatmap(process_name, heart_gene_sets[[process_name]])
}

all_target_ids <- target_presence |>
  dplyr::filter(present) |>
  dplyr::distinct(ensembl_id, .keep_all = TRUE) |>
  dplyr::pull(ensembl_id)

if (length(all_target_ids) >= 2) {
  mapped_all <- target_presence |>
    dplyr::filter(present) |>
    dplyr::distinct(ensembl_id, .keep_all = TRUE)

  labels_use <- mapped_all$symbol[match(all_target_ids, mapped_all$ensembl_id)]
  lfc_mat <- build_lfc_matrix(all_target_ids, row_labels = labels_use, cap = 2.5)
  padj_mat <- build_padj_matrix(all_target_ids, row_labels = labels_use)
  sig_mat <- make_sig_labels(padj_mat)
  row_category <- mapped_all$category[match(all_target_ids, mapped_all$ensembl_id)]
  row_category <- row_category[seq_len(nrow(lfc_mat))]

  ht <- ComplexHeatmap::Heatmap(
    lfc_mat,
    name = "log2FC",
    col = circlize::colorRamp2(c(-2.5, 0, 2.5), c("#2166AC", "white", "#B2182B")),
    na_col = "grey92",
    row_split = row_category,
    cluster_rows = TRUE,
    cluster_columns = FALSE,
    column_names_rot = 45,
    row_names_gp = grid::gpar(fontsize = 7, fontface = "italic"),
    column_title = "Mouse heart targeted biological panels - contrast-level log2FC",
    rect_gp = grid::gpar(col = "white", lwd = 0.4),
    cell_fun = function(j, i, x, y, width, height, fill) {
      if (sig_mat[i, j] != "") {
        grid::grid.text(sig_mat[i, j], x = x, y = y, gp = grid::gpar(fontsize = 7, fontface = "bold"))
      }
    },
    heatmap_legend_param = list(title = "log2FC")
  )

  fn_svg <- file.path(DIRS$targeted, "Heatmap_all_targeted_cardiac_panels_by_contrast_mouse_heart_diet.svg")
  fn_png <- file.path(DIRS$targeted, "Heatmap_all_targeted_cardiac_panels_by_contrast_mouse_heart_diet.png")
  grDevices::svg(fn_svg, width = 9, height = 16)
  ComplexHeatmap::draw(ht)
  grDevices::dev.off()
  png(fn_png, width = 9, height = 16, units = "in", res = 300)
  ComplexHeatmap::draw(ht)
  dev.off()
  ComplexHeatmap::draw(ht)

  readr::write_csv(
    tibble::rownames_to_column(as.data.frame(lfc_mat), "gene_symbol"),
    file.path(DIRS$targeted, "targeted_cardiac_panels_by_contrast_log2FC_matrix.csv")
  )
}

target_deg_table <- dplyr::bind_rows(lapply(names(contrast_results), function(nm) {
  contrast_results[[nm]] |>
    dplyr::filter(symbol_upper %in% unique(toupper(unlist(heart_gene_sets)))) |>
    dplyr::select(contrast, symbol, ensembl_id, log2FoldChange, stat, pvalue, padj, status, direction)
}))
readr::write_csv(target_deg_table, file.path(DIRS$targeted, "targeted_gene_DE_results_mouse_heart_diet.csv"))
```

## Boxplots for interesting cardiac genes

```{r gene-boxplots, fig.width=12, fig.height=8}
# --------------------------------------------------------------------------
# INDIVIDUAL GENE EXPRESSION BOXPLOTS
# VST-normalised expression is plotted for each gene in interesting_genes.
# Each dot represents one mouse (labelled with its ID); boxes show the
# interquartile range. Plots are saved individually and combined into
# paginated panels of 12 genes (4 columns × 3 rows) for the report.
# --------------------------------------------------------------------------
norm_counts <- DESeq2::counts(dds, normalized = TRUE)

make_gene_expression_table <- function(gene_symbols) {
  mapped <- map_gene_symbols(gene_symbols) |>
    dplyr::filter(present)

  if (nrow(mapped) == 0) return(tibble::tibble())

  dplyr::bind_rows(lapply(seq_len(nrow(mapped)), function(i) {
    gene_id <- mapped$ensembl_id[i]
    gene_symbol <- mapped$symbol[i]
    tibble::tibble(
      sample = colnames(vst_mat),
      ensembl_id = gene_id,
      symbol = gene_symbol,
      vst_expression = as.numeric(vst_mat[gene_id, ]),
      normalized_count = as.numeric(norm_counts[gene_id, ])
    ) |>
      dplyr::left_join(sample_meta |> dplyr::select(sample, condition_label, mouse_id), by = "sample")
  }))
}

gene_expression_long <- make_gene_expression_table(interesting_genes)
readr::write_csv(gene_expression_long, file.path(DIRS$boxplots, "interesting_gene_expression_mouse_heart_diet.csv"))

make_single_gene_boxplot <- function(df_gene) {
  gene_symbol <- unique(df_gene$symbol)[1]
  ggplot(df_gene, aes(condition_label, vst_expression, fill = condition_label)) +
    geom_boxplot(width = 0.62, outlier.shape = NA, alpha = 0.75, linewidth = 0.35) +
    geom_jitter(aes(label = mouse_id), width = 0.12, size = 2.4, alpha = 0.9) +
    ggrepel::geom_text_repel(aes(label = mouse_id), size = 2.6, max.overlaps = Inf, seed = 42, show.legend = FALSE) +
    scale_fill_manual(values = cond_colors, guide = "none") +
    labs(title = gene_symbol, x = NULL, y = "VST expression") +
    heart_theme(base_size = 10) +
    theme(axis.text.x = element_text(angle = 30, hjust = 1), plot.title = element_text(face = "italic"))
}

if (nrow(gene_expression_long) > 0) {
  gene_expression_split <- split(gene_expression_long, gene_expression_long$symbol)
  gene_boxplots <- lapply(gene_expression_split, make_single_gene_boxplot)

  for (gene_symbol in names(gene_boxplots)) {
    save_plot(gene_boxplots[[gene_symbol]], paste0("Boxplot_", safe_name(gene_symbol)), width = 4.2, height = 4.2, dir = DIRS$boxplots)
  }

  genes_per_page <- 12
  gene_names <- names(gene_boxplots)
  page_index <- split(gene_names, ceiling(seq_along(gene_names) / genes_per_page))
  for (i in seq_along(page_index)) {
    combined <- patchwork::wrap_plots(gene_boxplots[page_index[[i]]], ncol = 4)
    print(combined)
    save_plot(combined, paste0("Boxplots_interesting_genes_page_", i), width = 14, height = 10, dir = DIRS$boxplots)
  }
}
```

# VERTIENTE 4. Functional enrichment

## MSigDB gene sets and FGSEA

```{r fgsea, fig.width=9, fig.height=7}
# --------------------------------------------------------------------------
# FUNCTIONAL ENRICHMENT: FGSEA
# Gene set enrichment analysis is performed with fgsea on three MSigDB
# collections for each of the three contrasts (9 runs total):
#   - Hallmark (H): 50 curated hallmark gene sets
#   - GO Biological Process (C5 / GO:BP)
#   - KEGG pathways (C2 / CP:KEGG)
#
# The ranking metric is the DESeq2 Wald statistic (stat), which encodes
# both direction and magnitude of differential expression. When a gene
# symbol maps to multiple Ensembl IDs, only the entry with the highest
# |stat| is used (slice_max).
#
# fgsea parameters: minSize = 15, maxSize = 500 (standard for bulk RNA-seq).
# Results from all collections and contrasts are concatenated into fgsea_all
# and written to disk both as a single combined CSV and as per-contrast /
# per-collection CSVs.
#
# load_msigdb() handles API differences between msigdbr >= 7.5 (which uses
# collection/subcollection) and older versions (category/subcategory).
# --------------------------------------------------------------------------
load_msigdb <- function(collection, subcollection = NULL) {
  fml <- names(formals(msigdbr::msigdbr))
  out <- tryCatch({
    if ("collection" %in% fml) {
      if (!is.null(subcollection) && "subcollection" %in% fml) {
        msigdbr::msigdbr(species = "Mus musculus", collection = collection, subcollection = subcollection)
      } else {
        msigdbr::msigdbr(species = "Mus musculus", collection = collection)
      }
    } else {
      if (!is.null(subcollection)) {
        msigdbr::msigdbr(species = "Mus musculus", category = collection, subcategory = subcollection)
      } else {
        msigdbr::msigdbr(species = "Mus musculus", category = collection)
      }
    }
  }, error = function(e) {
    message("Could not load MSigDB collection ", collection, " / ", subcollection, ": ", conditionMessage(e))
    tibble::tibble()
  })
  out
}

prepare_pathways <- function(msig_tbl) {
  if (nrow(msig_tbl) == 0) return(list())
  msig_tbl_clean <- msig_tbl |>
    dplyr::filter(!is.na(gs_name), !is.na(gene_symbol)) |>
    dplyr::select(gs_name, gene_symbol) |>
    dplyr::distinct()
  split(msig_tbl_clean, msig_tbl_clean$gs_name) |>
    lapply(function(x) unique(x$gene_symbol))
}

msig_hallmark <- load_msigdb("H")
msig_go_bp <- load_msigdb("C5", "GO:BP")
msig_kegg <- load_msigdb("C2", "CP:KEGG")

pathway_collections <- list(
  Hallmark = prepare_pathways(msig_hallmark),
  GO_BP = prepare_pathways(msig_go_bp),
  KEGG = prepare_pathways(msig_kegg)
)
pathway_collections <- pathway_collections[lengths(pathway_collections) > 0]

prepare_ranked_stats <- function(res_df) {
  ranked <- res_df |>
    dplyr::filter(!is.na(symbol), !is.na(stat), !stringr::str_detect(symbol, "^ENSMUSG")) |>
    dplyr::group_by(symbol) |>
    dplyr::slice_max(order_by = abs(stat), n = 1, with_ties = FALSE) |>
    dplyr::ungroup() |>
    dplyr::arrange(dplyr::desc(stat))

  ranks <- ranked$stat
  names(ranks) <- ranked$symbol
  sort(ranks, decreasing = TRUE)
}

run_fgsea_one <- function(res_df, contrast_name, collection_name, pathways) {
  ranks <- prepare_ranked_stats(res_df)
  if (length(ranks) < 100 || length(pathways) == 0) return(tibble::tibble())

  fg <- fgsea::fgsea(
    pathways = pathways,
    stats = ranks,
    minSize = 15,
    maxSize = 500
  ) |>
    as_tibble() |>
    dplyr::arrange(padj, dplyr::desc(abs(NES))) |>
    dplyr::mutate(
      contrast = contrast_name,
      collection = collection_name,
      pathway_clean = clean_pathway_name(pathway)
    ) |>
    dplyr::relocate(contrast, collection)

  fg
}

fgsea_results <- list()
for (contrast_name in names(contrast_results)) {
  for (collection_name in names(pathway_collections)) {
    key <- paste(contrast_name, collection_name, sep = "__")
    fgsea_results[[key]] <- run_fgsea_one(
      contrast_results[[contrast_name]],
      contrast_name,
      collection_name,
      pathway_collections[[collection_name]]
    )
  }
}

fgsea_all <- dplyr::bind_rows(fgsea_results)

fgsea_all_csv <- fgsea_all |>
  dplyr::mutate(leadingEdge = vapply(leadingEdge, paste, collapse = ";", FUN.VALUE = character(1)))
readr::write_csv(fgsea_all_csv, file.path(DIRS$enrichment, "FGSEA_all_results_mouse_heart_diet.csv"))

for (contrast_name in unique(fgsea_all_csv$contrast)) {
  for (collection_name in unique(fgsea_all_csv$collection)) {
    tab <- fgsea_all_csv |>
      dplyr::filter(contrast == contrast_name, collection == collection_name)
    if (nrow(tab) > 0) {
      readr::write_csv(
        tab,
        file.path(DIRS$enrichment, paste0("FGSEA_", safe_name(collection_name), "_", safe_name(contrast_name), ".csv"))
      )
    }
  }
}

make_kable(
  fgsea_all_csv |>
    dplyr::filter(padj < PADJ_CUTOFF) |>
    dplyr::select(contrast, collection, pathway_clean, NES, pval, padj, size) |>
    dplyr::arrange(padj) |>
    dplyr::slice_head(n = 30),
  "Top significant FGSEA terms across contrasts"
)
```

## Enrichment dotplots

```{r enrichment-dotplots, fig.width=9, fig.height=7}
make_fgsea_dotplot <- function(fgsea_df, contrast_name, collection_name) {
  tab <- fgsea_df |>
    dplyr::filter(contrast == contrast_name, collection == collection_name, !is.na(padj), !is.na(NES))

  if (nrow(tab) == 0) return(invisible(NULL))

  plot_df <- dplyr::bind_rows(
    tab |> dplyr::filter(NES > 0) |> dplyr::arrange(padj) |> dplyr::slice_head(n = TOP_ENRICHMENT_TERMS / 2),
    tab |> dplyr::filter(NES < 0) |> dplyr::arrange(padj) |> dplyr::slice_head(n = TOP_ENRICHMENT_TERMS / 2)
  ) |>
    dplyr::distinct(pathway, .keep_all = TRUE) |>
    dplyr::mutate(
      pathway_clean = stringr::str_trunc(pathway_clean, 65),
      significant = padj < PADJ_CUTOFF
    )

  if (nrow(plot_df) == 0) return(invisible(NULL))

  p <- ggplot(plot_df, aes(NES, forcats::fct_reorder(pathway_clean, NES))) +
    geom_vline(xintercept = 0, linewidth = 0.35, color = "grey40") +
    geom_point(aes(size = -log10(padj), color = significant), alpha = 0.85) +
    scale_color_manual(values = c(`TRUE` = "#B2182B", `FALSE` = "grey55"), name = paste0("FDR < ", PADJ_CUTOFF)) +
    scale_size_continuous(name = expression(-log[10]~adjusted~italic(p))) +
    labs(
      title = paste(collection_name, "FGSEA -", contrast_name),
      x = "Normalized enrichment score (NES)",
      y = NULL
    ) +
    heart_theme()

  print(p)
  save_plot(p, paste0("FGSEA_dotplot_", safe_name(collection_name), "_", safe_name(contrast_name)), width = 9, height = 7, dir = DIRS$enrichment)
}

for (contrast_name in unique(fgsea_all$contrast)) {
  for (collection_name in unique(fgsea_all$collection)) {
    make_fgsea_dotplot(fgsea_all, contrast_name, collection_name)
  }
}
```

## GSEA enrichment plots

```{r gsea-enrichment-plots, fig.width=8, fig.height=4}
# --------------------------------------------------------------------------
# GSEA ENRICHMENT PLOTS (contrast-level)
# For each contrast × collection combination, the top N_GSEA_PLOTS pathways
# are visualised using fgsea::plotEnrichment(). The function generates the
# classic three-panel enrichment plot:
#   - Top panel:    running enrichment score curve (green line).
#   - Middle panel: rug plot marking where each gene set member falls in the
#                   ranked list (dense = early hits, sparse = late hits).
#   - Bottom panel: ranked list metric (DESeq2 Wald statistic) across all
#                   genes, from most upregulated (left) to most
#                   downregulated (right).
#
# Terms are selected separately for the positive and negative NES tails so
# that both directions are represented. The subtitle of each plot shows the
# contrast, collection, NES and FDR for traceability.
# Each plot is saved as SVG + PNG in DIRS$enrichment with the naming pattern:
#   GSEA_enrichplot_<collection>_<contrast>_<pathway>.svg/.png
# --------------------------------------------------------------------------
N_GSEA_PLOTS <- 3  # number of top terms per direction (up + down) per contrast

make_gsea_enrichment_plots <- function(fgsea_df, contrast_name, collection_name,
                                       pathways, n_top = N_GSEA_PLOTS) {
  tab <- fgsea_df |>
    dplyr::filter(
      contrast == contrast_name,
      collection == collection_name,
      !is.na(padj),
      padj < PADJ_CUTOFF,
      !is.na(NES)
    ) |>
    dplyr::arrange(padj, dplyr::desc(abs(NES)))

  top_terms <- dplyr::bind_rows(
    tab |> dplyr::filter(NES > 0) |> dplyr::slice_head(n = n_top),
    tab |> dplyr::filter(NES < 0) |> dplyr::slice_head(n = n_top)
  )

  if (nrow(top_terms) == 0) {
    message("No significant terms for GSEA enrichment plots: ", contrast_name, " - ", collection_name)
    return(invisible(NULL))
  }

  # Ranked stats vector (DESeq2 Wald statistic) for this contrast
  ranks <- prepare_ranked_stats(contrast_results[[contrast_name]])

  for (i in seq_len(nrow(top_terms))) {
    term      <- top_terms$pathway[i]
    term_clean <- top_terms$pathway_clean[i]

    if (!term %in% names(pathways)) next

    p <- fgsea::plotEnrichment(pathways[[term]], ranks) +
      ggplot2::labs(
        title    = stringr::str_trunc(term_clean, 70),
        subtitle = paste0(
          contrast_display_names[contrast_name], "  |  ", collection_name,
          "  |  NES = ", round(top_terms$NES[i], 2),
          "  |  FDR = ", formatC(top_terms$padj[i], format = "e", digits = 2)
        ),
        x = "Rank in ordered dataset",
        y = "Enrichment score"
      ) +
      heart_theme(base_size = 11)

    print(p)

    fn <- paste0(
      "GSEA_enrichplot_",
      safe_name(collection_name), "_",
      safe_name(contrast_name), "_",
      safe_name(term_clean)
    )
    save_plot(p, fn, width = 8, height = 4, dir = DIRS$enrichment)
  }
}

for (contrast_name in contrast_order) {
  for (collection_name in names(pathway_collections)) {
    make_gsea_enrichment_plots(fgsea_all, contrast_name, collection_name,
                               pathway_collections[[collection_name]])
  }
}
```

## Enrichment NES heatmaps

```{r enrichment-heatmaps, fig.width=10, fig.height=12}
make_nes_heatmap <- function(fgsea_df, filename, title, filter_regex = NULL, max_terms = TOP_ENRICHMENT_HEATMAP_TERMS) {
  tab <- fgsea_df |>
    dplyr::filter(
      contrast %in% contrast_order,
      !is.na(NES),
      !is.na(padj)
    ) |>
    dplyr::mutate(
      contrast = factor(contrast, levels = contrast_order),
      contrast_label = contrast_display_names[as.character(contrast)]
    )

  if (!is.null(filter_regex)) {
    tab <- tab |>
      dplyr::filter(stringr::str_detect(tolower(pathway_clean), stringr::regex(filter_regex, ignore_case = TRUE)))
  }

  if (nrow(tab) == 0) return(invisible(NULL))

  selected_terms <- tab |>
    dplyr::group_by(collection, pathway, pathway_clean) |>
    dplyr::summarise(
      min_padj = min(padj, na.rm = TRUE),
      max_abs_nes = max(abs(NES), na.rm = TRUE),
      significant_any = any(padj < PADJ_CUTOFF, na.rm = TRUE),
      .groups = "drop"
    ) |>
    dplyr::arrange(dplyr::desc(significant_any), min_padj, dplyr::desc(max_abs_nes)) |>
    dplyr::slice_head(n = max_terms)

  heat_df <- tab |>
    dplyr::semi_join(selected_terms, by = c("collection", "pathway", "pathway_clean")) |>
    dplyr::mutate(term = paste(collection, stringr::str_trunc(pathway_clean, 70), sep = " | ")) |>
    dplyr::select(term, contrast_label, NES, padj) |>
    dplyr::group_by(term, contrast_label) |>
    dplyr::summarise(NES = NES[which.min(padj)], padj = min(padj, na.rm = TRUE), .groups = "drop")

  if (nrow(heat_df) < 2) return(invisible(NULL))

  nes_mat <- heat_df |>
    dplyr::select(term, contrast_label, NES) |>
    tidyr::pivot_wider(names_from = contrast_label, values_from = NES) |>
    tibble::column_to_rownames("term") |>
    as.matrix()

  nes_mat <- nes_mat[, unname(contrast_display_names[contrast_order]), drop = FALSE]
  nes_mat[is.infinite(nes_mat)] <- NA_real_
  nes_mat_plot <- pmax(pmin(nes_mat, 3), -3)

  padj_mat <- heat_df |>
    dplyr::select(term, contrast_label, padj) |>
    tidyr::pivot_wider(names_from = contrast_label, values_from = padj) |>
    tibble::column_to_rownames("term") |>
    as.matrix()
  padj_mat <- padj_mat[rownames(nes_mat_plot), colnames(nes_mat_plot), drop = FALSE]
  sig_mat <- make_sig_labels(padj_mat)

  ht <- ComplexHeatmap::Heatmap(
    nes_mat_plot,
    name = "NES",
    col = circlize::colorRamp2(c(-3, 0, 3), c("#2166AC", "white", "#B2182B")),
    na_col = "grey92",
    cluster_rows = TRUE,
    cluster_columns = FALSE,
    row_names_gp = grid::gpar(fontsize = 7),
    column_names_rot = 45,
    column_title = title,
    rect_gp = grid::gpar(col = "white", lwd = 0.4),
    cell_fun = function(j, i, x, y, width, height, fill) {
      if (sig_mat[i, j] != "") {
        grid::grid.text(sig_mat[i, j], x = x, y = y, gp = grid::gpar(fontsize = 7, fontface = "bold"))
      }
    },
    heatmap_legend_param = list(title = "NES")
  )

  fn_svg <- file.path(DIRS$enrichment, paste0(filename, ".svg"))
  fn_png <- file.path(DIRS$enrichment, paste0(filename, ".png"))
  grDevices::svg(fn_svg, width = 9, height = max(6, 0.24 * nrow(nes_mat_plot) + 2))
  ComplexHeatmap::draw(ht)
  grDevices::dev.off()
  png(fn_png, width = 9, height = max(6, 0.24 * nrow(nes_mat_plot) + 2), units = "in", res = 300)
  ComplexHeatmap::draw(ht)
  dev.off()
  ComplexHeatmap::draw(ht)

  readr::write_csv(
    tibble::rownames_to_column(as.data.frame(nes_mat), "pathway"),
    file.path(DIRS$enrichment, paste0(filename, "_NES_matrix.csv"))
  )
}

make_nes_heatmap(
  fgsea_all,
  filename = "FGSEA_NES_heatmap_top_terms_by_contrast_mouse_heart_diet",
  title = "Top enriched pathways - contrast-level NES"
)

heart_relevant_regex <- paste(
  c(
    "heart", "cardiac", "myocard", "muscle", "contract", "calcium",
    "mitochond", "oxidative phosphorylation", "respiratory chain",
    "fatty acid", "lipid", "cholesterol", "peroxisome",
    "inflamm", "immune", "cytokine", "macrophage", "leukocyte",
    "fibro", "extracellular matrix", "collagen", "angiogenesis", "vascular", "endothelial",
    "oxidative stress", "reactive oxygen"
  ),
  collapse = "|"
)

make_nes_heatmap(
  fgsea_all,
  filename = "FGSEA_NES_heatmap_heart_relevant_processes_by_contrast_mouse_heart_diet",
  title = "Heart-relevant enriched processes - contrast-level NES",
  filter_regex = heart_relevant_regex,
  max_terms = 60
)
```

## Chord plots: enriched pathways and leading-edge genes

```{r chord-plots, fig.width=10, fig.height=10}
# --------------------------------------------------------------------------
# CHORD PLOTS: PATHWAYS AND LEADING-EDGE GENES
# Each chord plot links enriched pathways (left sectors) to their leading-edge
# genes (right sectors) for one contrast × collection combination.
# Colour convention: red = upregulated in the numerator of the contrast;
#                    blue = downregulated.
# Gene direction is taken from the DESeq2 log2FC of the same contrast.
# Pathway direction is taken from the sign of the FGSEA NES.
# A gene that appears in both up- and downregulated pathways is coloured grey.
# Up to max_terms pathways and max_genes_per_term leading-edge genes are shown
# to keep the plot readable. Chord plots are generated for Hallmark and GO_BP.
# --------------------------------------------------------------------------
make_chord_plot <- function(fgsea_df, contrast_name, collection_name, max_terms = 6, max_genes_per_term = 8) {
  tab <- fgsea_df |>
    dplyr::filter(
      contrast == contrast_name,
      collection == collection_name,
      !is.na(padj),
      padj < PADJ_CUTOFF,
      !is.na(NES)
    ) |>
    dplyr::arrange(padj, dplyr::desc(abs(NES))) |>
    dplyr::slice_head(n = max_terms)

  if (nrow(tab) == 0) {
    message("No significant terms for chord plot: ", contrast_name, " - ", collection_name)
    return(invisible(NULL))
  }

  # Gene direction is taken from the DESeq2 log2FC of the same contrast.
  # Red = upregulated in the numerator of the contrast; blue = downregulated.
  direction_tbl <- contrast_results[[contrast_name]] |>
    dplyr::filter(!is.na(symbol), !is.na(log2FoldChange)) |>
    dplyr::group_by(symbol) |>
    dplyr::slice_max(order_by = abs(stat), n = 1, with_ties = FALSE) |>
    dplyr::ungroup() |>
    dplyr::transmute(
      gene = symbol,
      log2FoldChange,
      direction = dplyr::case_when(
        log2FoldChange > 0 ~ "Up",
        log2FoldChange < 0 ~ "Down",
        TRUE ~ "Neutral"
      )
    )

  chord_df <- dplyr::bind_rows(lapply(seq_len(nrow(tab)), function(i) {
    genes <- tab$leadingEdge[[i]]
    genes <- genes[!is.na(genes)]
    genes <- head(genes, max_genes_per_term)
    if (length(genes) == 0) return(tibble::tibble())
    tibble::tibble(
      pathway = stringr::str_trunc(tab$pathway_clean[i], 36),
      gene = genes,
      NES = tab$NES[i],
      pathway_direction = dplyr::case_when(
        tab$NES[i] > 0 ~ "Up",
        tab$NES[i] < 0 ~ "Down",
        TRUE ~ "Neutral"
      ),
      value = 1
    )
  })) |>
    dplyr::left_join(direction_tbl, by = "gene") |>
    dplyr::mutate(
      direction = dplyr::coalesce(direction, pathway_direction, "Neutral"),
      link_color = dplyr::case_when(
        direction == "Up" ~ up_down_colors["Up"],
        direction == "Down" ~ up_down_colors["Down"],
        TRUE ~ up_down_colors["Neutral"]
      )
    )

  if (nrow(chord_df) == 0) return(invisible(NULL))

  chord_links <- chord_df |>
    dplyr::select(pathway, gene, value)

  pathway_colors <- chord_df |>
    dplyr::distinct(pathway, pathway_direction) |>
    dplyr::mutate(color = dplyr::case_when(
      pathway_direction == "Up" ~ up_down_colors["Up"],
      pathway_direction == "Down" ~ up_down_colors["Down"],
      TRUE ~ "grey75"
    ))

  gene_colors <- chord_df |>
    dplyr::group_by(gene) |>
    dplyr::summarise(
      direction = dplyr::case_when(
        any(direction == "Up") & !any(direction == "Down") ~ "Up",
        any(direction == "Down") & !any(direction == "Up") ~ "Down",
        any(direction == "Up") & any(direction == "Down") ~ "Neutral",
        TRUE ~ "Neutral"
      ),
      .groups = "drop"
    ) |>
    dplyr::mutate(color = dplyr::case_when(
      direction == "Up" ~ up_down_colors["Up"],
      direction == "Down" ~ up_down_colors["Down"],
      TRUE ~ up_down_colors["Neutral"]
    ))

  grid_col <- c(
    stats::setNames(pathway_colors$color, pathway_colors$pathway),
    stats::setNames(gene_colors$color, gene_colors$gene)
  )
  link_col <- chord_df$link_color

  plot_chord_once <- function() {
    circlize::circos.clear()
    circlize::chordDiagram(
      chord_links,
      col = link_col,
      grid.col = grid_col,
      transparency = 0.20,
      annotationTrack = "grid",
      preAllocateTracks = 1
    )
    circlize::circos.trackPlotRegion(
      track.index = 1,
      panel.fun = function(x, y) {
        sector.name <- circlize::get.cell.meta.data("sector.index")
        xlim <- circlize::get.cell.meta.data("xlim")
        ylim <- circlize::get.cell.meta.data("ylim")
        circlize::circos.text(
          mean(xlim), ylim[1] + 0.1, sector.name,
          facing = "clockwise", niceFacing = TRUE,
          adj = c(0, 0.5), cex = 0.45
        )
      },
      bg.border = NA
    )
    title(paste(collection_name, "-", contrast_display_names[contrast_name], "\nred = upregulated; blue = downregulated"))
    legend(
      "bottomleft",
      legend = c("Upregulated", "Downregulated"),
      fill = c(up_down_colors["Up"], up_down_colors["Down"]),
      border = NA,
      bty = "n",
      cex = 0.8
    )
    circlize::circos.clear()
  }

  fn_svg <- file.path(DIRS$enrichment, paste0("Chord_up_down_", safe_name(collection_name), "_", safe_name(contrast_name), ".svg"))
  fn_png <- file.path(DIRS$enrichment, paste0("Chord_up_down_", safe_name(collection_name), "_", safe_name(contrast_name), ".png"))

  grDevices::svg(fn_svg, width = 10, height = 10)
  plot_chord_once()
  grDevices::dev.off()

  png(fn_png, width = 10, height = 10, units = "in", res = 300)
  plot_chord_once()
  dev.off()

  plot_chord_once()

  readr::write_csv(
    chord_df |>
      dplyr::select(pathway, gene, NES, pathway_direction, log2FoldChange, direction),
    file.path(DIRS$enrichment, paste0("Chord_table_up_down_", safe_name(collection_name), "_", safe_name(contrast_name), ".csv"))
  )
}

for (contrast_name in contrast_order) {
  for (collection_name in intersect(c("Hallmark", "GO_BP"), unique(fgsea_all$collection))) {
    make_chord_plot(fgsea_all, contrast_name, collection_name)
  }
}
```

# Output and reproducibility

```{r output-summary}
generated_files <- list.files(OUT_DIR, recursive = TRUE, full.names = TRUE)
cat("Generated files:", length(generated_files), "\n")
data.frame(file = generated_files) |> make_kable("Generated mouse heart diet output files")
```

```{r session-info}
sessionInfo()
```
